Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-11-13 paper/appendix.tex
clarification thresholding

\subsection{Binarizing $\hat{H}$ - the Sonquist-Morgan criterion}

Back to back regression provides us with $Diag(\hat{H})$, a vector of real numbers that estimate the causal influence of features in $X$. In order to derive the true causal influence matrix $E$ , a binary matrix, we need to binarize  $\hat{H}$, i.e. test each feature for equality with zero. In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}. This test will not succeed here: $\hat{H}$ is obtained from a double regression with regularization, and will not follow the t-distribution.

Instead, we treat the binarization of $\hat{H}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''. More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} \citep{kass75} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{H}$, set to one all the $p$ diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the $d_x-p$ diagonal entries.

>>>>

\subsection{Recovering E}

Three distinct thresholding rules can be used to binarize $\text{diag(\hat H)}$ and tbus explicitly recover causal features under noise. (In the MEG experiment, we implement the third one).

First, given known signal-to-noise ratio, one can derive the bound above which a feature should considered to be causal. Indeed, Equation $\ref:diagk$ implies that the $k$ first diagonal elements of $\hat H$ are bounded:  % we assume that $F$ has full rank over $Img(E)$.

$$0 \leq \frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq \text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
where $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$.

The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$. In practive $\mu$ can be estimated as :
\begin{equation}
\mu = \frac{Var(X)}{Var(X)+Var(N)}
\end{equation}

The decision threshold between causal and non-causal elements is thus a fraction $\mu$. This threshold is to be adjusted depending on the type I and type II risks.

Second, $\text{diag{\hat H}}$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a clustering procedure separating small and large values. This procedure maximizes the ratio of inter-group variance ands minimizes the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan criterion maximizes \citep{sonquist_morgan} \citep{kass75}:
\begin{equation}
  \frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}
\end{equation}
%
Finally, second-order statistics across multiple datasets can be used to identify the elements of $\text{diag}(\hat H)$ that are sgnificantly different from 0. This procedure is detailed in the methods.

### 2019-11-13 paper/paper.tex
clarification thresholding

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X_1 X_1} & \Sigma_{X_1 X_2} \\ \Sigma_{X_1 X_2} & \Sigma_{X_2 X_2}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N_1 N_1} & \Sigma_{N_1 N_2} \\ \Sigma_{N_1 N_2} & \Sigma_{N_2 N_2}\end{array}\right),
\end{equation}
% $$
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right),
\end{equation}
\begin{equation}
  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$
Otherwise the $k$ first diagonal elements of $\hat H$ are all positive, since we assume that $F$ has full rank over $Img(E)$. In addition, they are bounded:
$$\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}} \leq \text{diag}_k(\hat{H})\leq  \frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$$
where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X_1 X_1}$ and $\Sigma_{N_1 N_1}$. This allows us to recover $E$ by identifying the non-zero elements on the diagonal of $\hat{H}$.

In practical cases, non-linearities and sampling error will add noise to the diagonal elements of $\hat H$. We need to test whether diagonal elements are different from zero. The average value $\mu$ of non-zero coefficients of $\text{diag}(\hat H)$ is the trace of $\hat H$ divided by $k$. $\mu$ can be estimated as follow :
\begin{equation}
$\mu = \frac{Var(X)}{Var(X)+Var(N)}$
\end{equation}
This provides a method to identify causal features
%Since $X$ and $N$ are not correlated, $\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1}$ should have a low condition number, which means  should provide a good estimate of $\frac{1}{k}Tr(\hat H)$.

To test whether a feature of $X$ has causal influence. An alternative is the Sonquist-Morgan criterion presented in the appendix.

>>>>

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X_1 X_1} & \Sigma_{X_1 X_2} \\ \Sigma_{X_1 X_2} & \Sigma_{X_2 X_2}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N_1 N_1} & \Sigma_{N_1 N_2} \\ \Sigma_{N_1 N_2} & \Sigma_{N_2 N_2}\end{array}\right),
\end{equation}
% $$
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right),
\end{equation}
\begin{equation}
  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
  \label{eq:diagk}
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$ Therefore, we recover $E$ from $\hat H$.

Otherwise, with noise, the causal factors of $E$ correspond to the positive elements of $\test{diag}(\hatH)$. The methods to recover them are presented in Appendix XXXX$$$
